---
title       : Distributions
subtitle    : September 18, 2013
author      : Jason Bryer (jason@bryer.org)
job         : epsy530.bryer.org
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
--- 

```{r setup, echo=FALSE, results='hide', warning=FALSE}
require(ggplot2)
require(gdata)
options(digits=4, width=95)
opts_chunk$set(fig.path='Figures/Class05-')
opts_chunk$set(comment=NA)
source('../R/contingency.table.R')


load('../Data/titanic.Rda')
movies <- read.csv('../Data/Textbook/Chapter_3/movie_lengths_2010.csv', stringsAsFactors=FALSE)
ozone <- read.csv('../Data/Textbook/Chapter_4/Ozone.csv', stringsAsFactors=FALSE)
ozone$Ozone <- as.numeric(ozone$Ozone)
ozone <- ozone[!is.na(ozone$Ozone),]

# See http://data.giss.nasa.gov/gistemp/tabledata_v3/GLB.Ts+dSST.txt
temp <- read.table('../Data/GlobalTemp.txt', header=TRUE, strip.white=TRUE)
temp$means <- rowMeans(aggregate(temp[,c("DJF","MAM","JJA","SON")], by=list(temp$Year), FUN="mean")[,2:5])
temp$meansF <- temp$means / 100 * 1.8

# temp2 <- melt(temp[,1:13], id='Year')
# temp2$value <- temp2$value / 100 * 1.8
# ggplot(temp2, aes(x=variable, y=value)) + geom_boxplot()

# World data
countries = read.csv('../Data/WDI_GDF_Country.csv', strip.white=TRUE)
worldData = read.csv('../Data/WDI_GDF_Data.csv', strip.white=TRUE)
worldData2 = worldData[which(worldData$Series.Code %in% c('NY.GDP.MKTP.KD', 'SE.XPD.TOTL.GD.ZS', 'SP.DYN.LE00.IN', 'SP.POP.TOTL')), c('Series.Code', 'Series.Name', 'Country.Name', 'Country.Code', 'X2008')]
worldData2 = merge(worldData2, countries[,c('Country.Code', 'Region')], by='Country.Code')
worldData2 = worldData2[which(worldData2$Region != 'Aggregates'),]
worldData2$Series.Name = as.factor(as.character(worldData2$Series.Name))
worldData2$Region = as.factor(as.character(worldData2$Region))
worldData3 = cast(worldData2, Country.Name + Region ~ Series.Name, mean, value='X2008')
names(worldData3) = c('Country', 'Region', 'GDP', 'Life.Expectancy', 'Population', 'Education')
worldData3$GDP.log = log(worldData3$GDP)
worldData3$GDP = worldData3$GDP / 1000000000 #Billions
worldData3$Population = worldData3$Population / 1000000 #Millions

```

## Comparing Histograms

```{r, warning=FALSE, fig.width=10, fig.height=6,tidy=FALSE, message=FALSE}
ggplot(titanic, aes(x=age)) + geom_histogram() + facet_wrap(~ pclass, ncol=1)
```

--- &twocol

## Timeplots

*** left
* Timeplots display every data value on a timeline.
* Great for spotting trends

*** right

```{r, warning=FALSE, fig.width=6, fig.height=5,tidy=FALSE}
ggplot(temp, aes(x=Year, y=meansF)) + 
	geom_point()
```

--- &twocol

## Connecting the Dots

*** left

* Connecting the dots of a timeplot can sometimes better illustrate the trends.
* This example has so many dots that this graph is busy and not that illustrative.
* Connecting the dots is better for either fewer data values or data with less variation.

*** right

```{r, warning=FALSE, fig.width=6, fig.height=5,tidy=FALSE}
ggplot(temp, aes(x=Year, y=meansF)) + 
	geom_point(alpha=.5) + 
	geom_line()
```

--- &twocol

## Smoothing the Data

*** left

* Drawing a curve of typical values in the neighborhood can sometimes tell the story better.
* There are many ways of doing this and a computer can be used to create this curve.
* The curve, called the lowess (or loess) curve, helps the eye follow the main trend and spot the outliers.


*** right

```{r, warning=FALSE, fig.width=6, fig.height=5, message=FALSE,tidy=FALSE}
ggplot(temp, aes(x=Year, y=meansF)) + 
	geom_point(alpha=.5) + 
	geom_line(alpha=.5) + 
	geom_smooth()
```

---

## Boxplots

```{r, warning=FALSE, fig.width=10, fig.height=5}
ggplot(ozone, aes(x=factor(Month), y=Ozone)) + geom_boxplot()
```

---

## Outliers

How to Approach Outliers

* Check to see if there may have been an error in the data collection or data input.
	* If the reported heights of students includes a student that is 170 inches tall (14 feet), maybe that student was measured in centimeters.
* Check to see if there was an extraordinary outcome.
	* The median number of daily customers at the Punxsutawney, PA, gift store may be 42 with an IQR of 12, but on February 2, there were 831 customers.

### Common Errors Causing an Outlier

* Transposing the digits
* A respondent not understanding the survey question
* Misreading results
* Confusion about units
* Cheating

---

## However, Outliers Can be the Most Interesting Data Values

* Income Data: The CEO
* Student Height: The basketball team’s center
* Snowfall: The great blizzard of ’98
* Exam Score:  The curve breaker
* Milk Purchased:  Octomom!

### Always comment on the outliers.

---

## Transforming Data

* When data is skewed it becomes difficult to interpret measures of center and spread.
* Transforming data is an approach to make skewed data more symmetric.

### Common Transformations

* Skewed Right: Use log, ln, or $\frac{1}{x}$
* Skewed Left: Use ${ x }^{ 2 }$


---

## Example: World GDP and Life Expectancy

```{r, warning=FALSE, fig.width=10, fig.height=5, message=FALSE,tidy=FALSE}
ggplot(worldData3, aes(x=GDP, y=Life.Expectancy)) + 
	geom_point(stat='identity', alpha=.6) + 
	xlab('Gross Domestic Product (billions)') + 
	ylab('Life Expectancy at birth (years)')
```

---

## Example: (log of) World GDP and Life Expectancy

```{r, warning=FALSE, fig.width=10, fig.height=5, message=FALSE,tidy=FALSE}
ggplot(worldData3, aes(x=log(GDP), y=Life.Expectancy)) + 
	geom_point(stat='identity', alpha=.6) + 
	xlab('Gross Domestic Product (billions)') + 
	ylab('Life Expectancy at birth (years)')
```

---

## On Comparing Distributions

### Choose the right tool.
* Use histograms to compare two or three groups.
* Use boxplots to compare many groups.

### Treat outliers with attention and care.
* Local or global, especially in a time series
* Investigate if the outliers are errors or remarkable.
* Use a timeplot to track trends over time.

### Re-express or transform data for better understanding.
* Can transform skewed distributions to symmetric ones
* Can help to compare spreads of different groups


