EPSY 530 Research Project - Prograamm of International Student Assessment
========================================================



[Click here to download the codebook](Codebook_Stu09_Dec10.pdf).


**NOTE:** The data used for this project has been simplified so inferences made with these data may not reflect what the full dataset may suggest. For example, any student records with any missing data were removed. Also, PISA uses a complex survey design where survey weights and multiple plausible values are provided for the outcome scores (i.e. math, reading, and science). The scores provided are simple means of the five possible values for simplicity.


### Load Data

```{r}
```


### Describe Variables

Question 28: Reading attitude
Question 29: How often do you read these materials because you want to?

```{r}
```


### Visualize Categorical Data

```{r}
```

### Visualize Quantitative Variables

```{r}
```


### Descriptive Statistics

Mean, median, standard deviation, interquartile range.

### Distributions

Describe the distrubition of the reading, math, and science scores.

```{r}
```


### Correlation

```{r}
```


### Linear Regression

```{r}
```


### Student t-tests

```{r}
```


